Modelling palaeoecological time series using generalized additive models

### 2018-08-16 manuscript.tex
bump date
minor change  to manuscript
insert flow diagram; bump date
mass rebuild after switching to gratia package

\date{May 11, 2018}

@@

Palaeoecology and palaeolimnology have moved away from being descriptive
disciplines, rapidly adopting new statistical developments in the 1990s
and beyond (@ Smol et al., 2012). Less development has been observed in
the area of trend estimation in palaeoenvironmental time series. The
vast majority of data produced by palaeoecologists and
palaeolimnologists is in the form of time-ordered observations on one or
more proxies or biological taxa (Birks, 2012b; Smol, 2008; Smol et al.,
2012). Typically these data are arranged irregularly in time; in the
absence of annual laminae or varves, the sediment core is sectioned at
regular depth intervals (Glew et al., 2001), which, owing to variation
in accumulation rates over time and compaction by overlying sediments,
results in an uneven sampling in time. An under-appreciated feature of
such sampling is that younger sediments often have larger variance than
older sediments; each section of core represents fewer lake years in
newer samples, relative to older samples. This variable averaging acts
as a time-varying low-pass (high-cut) filter of the annual depositional
signal.

Irregular intervals between samples means that the time-series analysis
methods of autoregressive or moving average processes, in the form of
autoregressive integrated moving average (ARIMA) models, are practically
impossible to apply because software typically requires even spacing of
observations in time. Dutilleul et al. (2012) and Birks (2012a),
eschewing the term \emph{time series}, prefer to call such data
\emph{temporal series} on account of the irregular spacing of samples, a
distinction that I find unnecessary.

Where statistical approaches have been applied to trend estimation in
palaeoenvironmental time series, a commonly-used method is \loess{}
(Birks, 1998, 2012a; Cleveland, 1979; Juggins and Telford, 2012).
\loess{}, locally weighted scatterplot smoother, as it's name suggests,
was developed to smooth x-y scatterplot data (Cleveland, 1979). The
method fits a smooth line through data by fitting weighted least squares
(WLS) models to observations within a particular, user-specified window
of the focal point, whose width is typically expressed as a proportion
\(\alpha\) of the \(n\) data points. Weights are determined by how close
(in the x-axis only) an observation in the window is to the focal point
giving greatest weight given to points closest to the focal point. The
interim \loess{}-smoothed value for the focal point is the predicted
value from the weighted regression at the focal point. The interim
values are updated using weights based on how far in the y-axis
direction the interim smoothed value lies from the observed value plus
the x-axis distance weights; this has the effect of down-weighting
outlier observations. The final \loess{} is obtained by joining the
smoothed values. The user has to choose how large a window to use,
whether to fit degree 1 (linear) or degree 2 (quadratic) polynomials in
the WLS model, and how to weight points in the x-axis. When used in an
exploratory mode, the user has considerable freedom to choose the detail
of the \loess{} fit; the window width, for example, can be infinitely
tweaked to give as close a fit to the data, as assessed by eye, as is
desired. Using cross-validation (CV) to choose \(\alpha\) or the degree
of polynomial in the WLS model is complicated for a number of reasons,
not least because the CV scheme used must involve the time ordering of
the data (e.g. Bergmeir et al., 2018). This subjectivity is problematic
however once we wish to move beyond exploratory analysis and
statistically identify trends to test hypotheses involving those trend
estimates.

@@

In the subsequent sections, I present an introduction to GAMs and
discuss the issue of uncertainty in model-estimated trends, the topic of
posterior simulation from a regression model and how to identify periods
of significant environmental change using the first derivative of the
estimated trend. Two non-standard types of spline --- adaptive smoothers
and Gaussian process splines --- that are especially applicable to GAMs
in the palaeoenvironmental setting are subsequently described, followed
by an assessment of the the impact of age-model uncertainty on trend
estimation via GAMs. Finally, I briefly discuss the application of GAM
trend analysis to multivariate species abundance and compositional data.

@@

To illustrate trend estimation in palaeoenvironmental data using GAMs, I
use two proxy time series; a 150-year bulk organic matter
δ\textsuperscript{15}N record from Small Water, and a 3000-year alkenone
record from Braya-Sø. Between them, the two examples, combine many of
the features of interest to palaeoecologists that motivate the use of
GAMs; non-linear trends and the question of when changes in the measured
proxy occurred. The example analyses were all performed using the
\emph{mgcv} package (version 1.8.23; Wood, 2017) and R (version 3.4.4; R
Core Team, 2018), and the supplementary material contains a fully
annotated document showing the R code used to replicate all the analyses
described in the remainder of the paper.

@@

The second example time series is a 3,000 year record of alkenone
unsaturation, \uk, from Braya-Sø, a meromictic lake in West Greenland
(D'Andrea et al., 2011). Alkenones are long-chained unsaturated organic
compounds that are produced by a small number of planktonic organisms
known as haptophytes. The \uk{} unsaturation index is (Brassell, 1993)

@@

It is clear from the plots of the data (Figure~\ref{fig:data-figure})
that we require the fitted trends for the Small Water
δ\textsuperscript{15}N and Braya-Sø \uk{} time series to be non-linear
functions, but it is less clear how to specify the actual shape require.
Ideally, we'd like to learn the shape of the trend from the data
themselves. We will refer to these non-linear functions as \emph{smooth
functions}, or just \emph{smooths} for short, and we will denote a
smooth using \(f(x_t)\). Further, we would like to represent the smooths
in a way that \eqref{eq:gam} is represented parametrically so that it
can be estimate within the well-studied GLM framework. This is achieved
by representing the smooth using a \emph{basis}. A basis is a set of
functions that collectively span a space of smooths that, we hope,
contains the true \(f(x_t)\) or a close approximation to it. The
functions in the basis are known as \emph{basis functions}, and arise
from a \emph{basis expansion} of a covariate. Writing \(b_j(x_t)\) as
the \(j\)th basis function of \(x_t\), the smooth \(f(x_t)\) can be
represented as a weighted sum of basis functions

@@

The CRS can be parameterized in a number of different ways. One requires
a knot at each unique data value in \(x_t\), which is computationally
inefficient. Another way of specifying a CRS basis is to parameterize in
terms of the value of the spline at the knots. Typically in this
parametrization there are many fewer knots than unique data, with the
knots distributed evenly over the range of \(x_t\) or at the quantiles
of \(x_t\). Placing knots at the quantiles of \(x_t\) has the effect of
placing a greater number of knots where the data is densest.

@@

To estimate a model using this basis expansion each basis function forms
a column in the model matrix \textbf{X} and the weights \(\beta_j\) can
be found using least squares regression (assuming a Gaussian response).
Note that in order to estimate a coefficient for each basis function the
model has to be fitted without an intercept term. In practice we would
include an intercept term in the model and therefore the basis functions
are modified via an identifiability constraint (@ Wood, 2017). This has
the effect of making the basis orthogonal to the intercept but results
in more complicated basis functions than those shown in in Figure
\ref{fig:basis-function-example-plot}a.

Having estimated the weight for each basis function, the \(j\)th basis
function \(b_j\) is scaled (weighted) by its coefficient \(\beta_j\).
The scaled CRS basis functions for the Small Water time series are shown
in Figure \ref{fig:basis-function-example-plot}b. The solid line passing
through the data points is formed by summing up the values of the scaled
basis functions (\(b_j(x_t) \beta_j\)) at any value of \(x_t\) (time).

@@

To fit a GAM using either of the two regression spline bases described
above, the analyst is generally only required to the specify the size
(rank) of the basis expansion required to represent or closely
approximate the true function \(f\). With practice and some knowledge of
the system from which the observations arise, it can be relatively easy
to put an upper limit on the expected complexity of the true trend in
the data. Additionally, the number available data points places a
constraint on the upper limit of the size of basis expansion that can be
used.

In practice, the size of the basis is an upper limit on the expected
complexity of the trend, and a simple test to check if the basis used
was sufficiently large (Pya and Wood, 2016). This test is available via
the \texttt{gam.check()} function in \emph{mgcv} for example, which
essentially looks at whether there is any additional nonlinearity or
structure in the residuals that can be explained by a further smooth of
\(x_t\). Should a smooth term in the fitted model fail this test the
model can be refitted using a larger basis expansion, say by doubling
the value of \emph{k} (the rank) used to fit the original. Note also
that a smooth might fail this test whilst using fewer effective degrees
of freedom than the maximum possible for the dimension of basis used.
This may happen when the true function lies at the upper limit of the
set of functions encompassed by the size of basis used. Additionally, a
basis of size \(2k\) encompasses a richer space of functions of a given
complexity than a basis of size \(k\) (Wood, 2017); increasing the basis
dimension used to fit the model may unlock this additional function
space resulting in a better fitting model whilst using a similar number
of effective degrees of freedom.

@@

With the larger basis dimension and accounting for the non-constant
variance of the observations via weights, the model fitted using REML is
indistinguishable from that obtained using GCV (Figure
\ref{fig:plot-fitted-models}b). The trace of the REML score for this
model shows a pronounced minimum at a much smaller value of \(\lambda\)
than the original REML fit (Figure
\ref{fig:trace-smoothness-parameters}), indicating that a more wiggly
trend provides a better fit to the Braya-Sø time series. This example
illustrates that some care and understanding of the underlying
principles of GAMs is required to diagnose potential issues with the
estimated model. After standard modelling choices (size of basis to use,
correct selection of response distribution and link function, etc.) are
checked, it often pays to think carefully about the properties of the
data and ensure that the assumptions of the model are met. Here, despite
increasing the basis size, it was the failure to appreciate the
magnitude of the effect on fitting of the non-constant variance that
lead to the initially poor fit and the problems associated with the
estimation of the CAR(1) process. I return to the issue of why the GAM
plus CAR(1) model encountered problems during fitting later
(\protect\hyperlink{identifiability}{see Identifiability}).

@@

If we want to ask whether either of the estimated trends is
statistically interesting or proceed to identifying periods of
significant change, we must address the issue of uncertainty in the
estimated model. What uncertainty is associated with the trend
estimates? One way to visualise this is through a 1 - \(\alpha\)
confidence interval around the fitted trend, where \(\alpha\) is say
0.05 leading to a 95\% interval. A 95\% interval would be drawn at
\(\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))\), with
\(m_{1-\alpha} = 1.96\), the 0.95 probability quantile of a standard
normal distribution, and \(\text{SE}(\hat{y}_t)\) is the standard error
of the estimated trend at time \(x_t\). This type of confidence interval
would normally be described as \emph{pointwise}; the coverage properties
of the interval being correct for a single point on the fitted trend,
but, if we were to consider additional points on the trend, the coverage
would logically be lower than 1 - \(\alpha\). This is the traditional
frequentist interpretation of a confidence interval. However, the GAM
described here has a Bayesian interpretation (Kimeldorf and Wahba, 1970;
Silverman, 1985; Wahba, 1983, 1990) and therefore the typical
frequentist interpretation does not apply. Nychka (1988) investigated
the properties of a confidence interval created as described above using
standard errors derived from the Bayesian posterior covariance matrix
for the estimated mode parameters. Such intervals have the interesting
property that they have good \emph{across-the-function} coverage when
considered from a frequentist perspective. This means that, when
averaged over the range of the function, the Bayesian credible intervals
shown in Figure \ref{fig:plot-fitted-models} have close to the expected
95\% coverage. However, to achieve this some parts of the function may
have more or less than 95\%-coverage. Marra and Wood (2012) recently
explained Nychka's (1988) surprising results and extended them to the
case of generalized models (non-Gaussian responses).

@@

Figure \ref{fig:posterior-simulation} shows 20 random draws from the
posterior distribution of the GAMs fitted to the Small Water and
Braya-Sø data sets. In the early period of the δ\textsuperscript{15}N
time series many of the posterior simulations exhibit short periods of
increasing and decreasing trend, balancing out to the relatively flat
trend estimated by the GAM (Fig.~\ref{fig:posterior-simulation}a).
Reflecting this uncertainty, we might expect relatively wide
simultaneous intervals during this period in order to contain the vast
majority of the simulated trends. Conversely, the decreasing
δ\textsuperscript{15}N trend starting at \textasciitilde{}1945 is
consistently reproduced in the posterior simulations, suggesting that
this feature of the time series is both real and statistically
significant, and that the rate of change in δ\textsuperscript{15}N is
relatively precisely estimated. We see a similar pattern in
Figure~\ref{fig:posterior-simulation}b for the Braya-Sø record; the
large peak in \uk{} at \textasciitilde{}250CE and the strong decline at
\textasciitilde{}1200CE are well defined in the posterior simulations,
whereas most of the localised trends that are smaller magnitude changes
in \(y_t\) are associated with posterior simulations that are less well
constrained with the ends of the record in particular showing
considerable variation in the strength, timing, and even sign of
simulated trends, reflecting the greater uncertainty in estimated trend
during these periods. For the random draws illustrated in Figure
\ref{fig:posterior-simulation}, a (1 - \(\alpha\))100\% simultaneous
interval should contain the entire function for on average 19 of the 20
draws.

@@

Simultaneous intervals computed using the method described are show in
Figure \ref{fig:compare-intervals} alongside the
\emph{across-the-function} confidence intervals for the trends fitted to
both example data sets. As expected, the simultaneous interval is
somewhat wider than the \emph{across-the-function} interval. The
critical value \(m_{1-\alpha}\) for the simultaneous interval of the
estimated trend in δ\textsuperscript{15}N is 3.08, whilst the same value
for the \uk{} series is 3.42, leading to intervals that are
approximately ±50\% and ±75\% wider than the equivalent
across-the-function intervals.

@@

Why might this be so? Autocorrelation is the tendency for a large
(small) value of \(y_t\) at time \(x_t\) to be followed by a likewise
large (small) value at time \(x_{t+1}\). This leads to runs of values
that are consistently greater (less) than the overall mean. Short runs
would indicate weaker autocorrelation whilst longer runs are associated
with stronger autocorrelation, and long runs of values greater (less)
than the mean would be evident as non-linear trends in the time series.
As a result, a wiggly trend and an autocorrelation function with large
\(\phi\) are two ways to describe the same pattern of values in a time
series, and without any further information to constrain either the
model is unable to distinguish both components uniquely.

@@

In the world of machine learning, Gaussian processes (Golding and Purse,
2016; Rasmussen and Williams, 2006) are a widely-used method for fitting
smooth non-parametric regression models. A Gaussian process is a
distribution over all possible smooth functions \(f(x)\). In the field
of spatial statistics, Gaussian processes are known by name
\emph{kriging}.

@@

Each of the spline types that I have discussed so far share a common
feature; the degree of wiggliness over the time series is fixed due to
the use of a single smoothness parameter, \(\lambda\). The definition of
wiggliness, as the integrated squared second derivative of the spline,
ensures that the fitted smoother does not jump about wildly. This
assumes that the data themselves are well described by a smoothly
varying trend. If we anticipate abrupt change or step-like responses to
environmental forcing this underlying assumption of the GAM would
suggest that the method is ill-suited to modelling palaeo time series in
which such features are evident or expected.

@@

Thus far, the trend models that I have described and illustrated assumed
that the time covariate (\(x_t\)) was fixed and known. In both examples,
and more generally for most palaeoecological records, this assumption is
violated. Unless the record is annually laminated, assigning an age to a
sediment interval requires the development of an age model from
observations of the relationship between depth down the sediment core
and estimates of the age of the sample arrived at using any of a number
of techniques, for example \textsuperscript{210}Pb or
\textsuperscript{14}C radiometric dating. This age-depth relationship is
itself uncertain, usually being derived from a mathematical or
statistical model applied to point age estimates (e.g.~ Blaauw and
Heegaard, 2012). Incorporating this additional component of uncertainty
complicates the estimation of statistical models from
palaeoenvironmental data. In this section I illustrate a simulation
based approach to quantify and account for age-model uncertainty as part
of the trend estimation using a GAM (see Anchukaitis and Tierney (2013)
for a similar, non-GAM related idea).

@@

The ideas expressed in this paper are the result of many fruitful
conversations with colleagues at the Environmental Change Research
Centre, UCL, and the University of Regina. In particular I am indebted
to Helen Bennion, Rick Battarbee, and Peter Leavitt for their
collaborations on projects over many years, and to David Miller, Eric
Pedersen, and Noam Ross, my GAM workshop partners in crime. Without
Simon Wood's \emph{mgcv} software and his research on GAMs, the
application of these models to palaeo time series would not be as
straight forward. This work was supported by a Natural Sciences and
Engineering Council of Canada (NSERC) Discovery Grant to the author
(RGPIN-2014-04032).

>>>>

\date{August 16, 2018}

@@

Palaeoecology and palaeolimnology have moved away from being descriptive
disciplines, rapidly adopting new statistical developments in the 1990s
and beyond (Smol et al., 2012). Less development has been observed in
the area of trend estimation in palaeoenvironmental time series. The
vast majority of data produced by palaeoecologists and
palaeolimnologists is in the form of time-ordered observations on one or
more proxies or biological taxa (Birks, 2012b; Smol, 2008; Smol et al.,
2012). Typically these data are arranged irregularly in time; in the
absence of annual laminae or varves, the sediment core is sectioned at
regular depth intervals (Glew et al., 2001), which, owing to variation
in accumulation rates over time and compaction by overlying sediments,
results in an uneven sampling in time. An under-appreciated feature of
such sampling is that younger sediments often have larger variance than
older sediments; each section of core represents fewer lake years in
newer samples, relative to older samples. This variable averaging acts
as a time-varying low-pass (high-cut) filter of the annual depositional
signal.

Irregular intervals between samples means that the time-series analysis
methods of autoregressive or moving average processes, in the form of
autoregressive integrated moving average (ARIMA) models, are practically
impossible to apply because software typically requires even spacing of
observations in time. Dutilleul et al. (2012) and Birks (2012a),
eschewing the term \emph{time series}, prefer to call such data
\emph{temporal series} on account of the irregular spacing of samples, a
distinction that I find unnecessary, however.

Where statistical approaches have been applied to trend estimation in
palaeoenvironmental time series, a commonly-used method is \loess{}
(Birks, 1998, 2012a; Cleveland, 1979; Juggins and Telford, 2012).
\loess{}, locally weighted scatterplot smoother, as it's name suggests,
was developed to smooth x-y scatterplot data (Cleveland, 1979). The
method fits a smooth line through data by fitting weighted least squares
(WLS) models to observations within a user-specified window of the focal
point, whose width is typically expressed as a proportion \(\alpha\) of
the \(n\) data points. Weights are determined by how close (in the
x-axis only) an observation in the window is to the focal point giving
greatest weight given to points closest to the focal point. The interim
\loess{}-smoothed value for the focal point is the predicted value from
the weighted regression at the focal point. The interim values are
updated using weights based on how far in the y-axis direction the
interim smoothed value lies from the observed value plus the x-axis
distance weights; this has the effect of down-weighting outlier
observations. The final \loess{} is obtained by joining the smoothed
values. The user has to choose how large a window to use, whether to fit
degree 1 (linear) or degree 2 (quadratic) polynomials in the WLS model,
and how to weight points in the x-axis. When used in an exploratory
mode, the user has considerable freedom to choose the detail of the
\loess{} fit; the window width, for example, can be infinitely tweaked
to give as close a fit to the data, as assessed by eye, as is desired.
Using cross-validation (CV) to choose \(\alpha\) or the degree of
polynomial in the WLS model is complicated for a number of reasons, not
least because the CV scheme used must involve the time ordering of the
data (e.g. Bergmeir et al., 2018). This subjectivity is problematic
however once we wish to move beyond exploratory analysis and
statistically identify trends to test hypotheses involving those trend
estimates.

@@

In the subsequent sections, I present an introduction to GAMs and
discuss the issue of uncertainty in model-estimated trends, the topic of
posterior simulation from a regression model and how to identify periods
of significant environmental change using the first derivative of the
estimated trend. The main steps in the analysis of palaeoenvironmental
time series using GAMs are illustrated in Figure \ref{fig:flowchart}.
Two non-standard types of spline --- adaptive smoothers and Gaussian
process splines --- that are especially applicable to GAMs in the
palaeoenvironmental setting are subsequently described, followed by an
assessment of the the impact of age-model uncertainty on trend
estimation via GAMs. Finally, I briefly discuss the application of GAM
trend analysis to multivariate species abundance and compositional data.

{\centering \includegraphics[width=0.4\linewidth]{manual-figures/flow-diagram}

\caption{Flowchart showing the main steps in the analysis of time series using generalized additive models.}\label{fig:flowchart}
\end{figure}

@@

To illustrate trend estimation in palaeoenvironmental data using GAMs, I
use two proxy time series; a 150-year bulk organic matter
δ\textsuperscript{15}N record from Small Water, and a 3000-year alkenone
record from Braya-Sø. Between them, the two examples, combine many of
the features of interest to palaeoecologists that motivate the use of
GAMs; non-linear trends and the question of when changes in the measured
proxy occurred. The example analyses were all performed using the
\emph{mgcv} package (version 1.8.24; Wood, 2017) and R (version 3.4.4; R
Core Team, 2018), and the supplementary material contains a fully
annotated document showing the R code used to replicate all the analyses
described in the remainder of the paper.

@@

The second example time series is a 3,000 year record of alkenone
unsaturation, \uk, from Braya-Sø, a meromictic lake in West Greenland
(D'Andrea et al., 2011). Alkenones are long-chained unsaturated organic
compounds that are produced by a small number of planktonic organisms
known as haptophytes. The \uk{} unsaturation index (Brassell, 1993) is

@@

It is clear from plots of the data (Figure~\ref{fig:data-figure}) that
we require the fitted trends for the Small Water δ\textsuperscript{15}N
and Braya-Sø \uk{} time series to be non-linear functions, but it is
less clear how to specify the actual shape require. Ideally, we'd like
to learn the shape of the trend from the data themselves. We will refer
to these non-linear functions as \emph{smooth functions}, or just
\emph{smooths} for short, and we will denote a smooth using \(f(x_t)\).
Further, we would like to represent the smooths in a way that
\eqref{eq:gam} is represented parametrically so that it can be estimate
within the well-studied GLM framework. This is achieved by representing
the smooth using a \emph{basis}. A basis is a set of functions that
collectively span a space of smooths that, we hope, contains the true
\(f(x_t)\) or a close approximation to it. The functions in the basis
are known as \emph{basis functions}, and arise from a \emph{basis
expansion} of a covariate. Writing \(b_j(x_t)\) as the \(j\)th basis
function of \(x_t\), the smooth \(f(x_t)\) can be represented as a
weighted sum of basis functions

@@

The CRS can be parameterized in a number of different ways. One requires
a knot at each unique data value in \(x_t\), which is computationally
inefficient. Another way of specifying a CRS basis is to parameterize in
terms of the value of the spline at the knots. Typically in this
parametrization there are many fewer knots than unique data, with the
knots distributed evenly over the range of \(x_t\) or at the quantiles
of \(x_t\). Placing knots at the quantiles of \(x_t\) has the effect of
placing a greater number of knots where the data is most dense.

@@

To estimate a model using this basis expansion each basis function forms
a column in the model matrix \textbf{X} and the weights \(\beta_j\) can
be found using least squares regression (assuming a Gaussian response).
Note that in order to estimate a coefficient for each basis function the
model has to be fitted without an intercept term. In practice we would
include an intercept term in the model and therefore the basis functions
are modified via an identifiability constraint (Wood, 2017). This has
the effect of making the basis orthogonal to the intercept but results
in more complicated basis functions than those shown in in Figure
\ref{fig:basis-function-example-plot}a.

Having estimated the weight for each basis function, the \(j\)th basis
function \(b_j\) is scaled (weighted) by its coefficient \(\beta_j\).
The scaled CRS basis functions for the Small Water time series are shown
in Figure \ref{fig:basis-function-example-plot}b. The solid line passing
through the data points is formed by summing up the values of the seven
scaled basis functions (\(b_j(x_t) \beta_j\)) at any value of \(x_t\)
(time).

@@

To fit a GAM using either of the two regression spline bases described
above, the analyst is generally only required to the specify the size
(rank) of the basis expansion required to represent or closely
approximate the true function \(f\). With practice and some knowledge of
the system from which the observations arise, it can be relatively easy
to put an upper limit on the expected complexity of the true trend in
the data. Additionally, the number of available data points places a
constraint on the upper limit of the size of basis expansion that can be
used.

In practice, the size of the basis is an upper limit on the expected
complexity of the trend, and a simple test can be used to check if the
basis used was sufficiently large (Pya and Wood, 2016). This test is
available via the \texttt{gam.check()} function in \emph{mgcv} for
example, which essentially looks at whether there is any additional
nonlinearity or structure in the residuals that can be explained by a
further smooth of \(x_t\). Should a smooth term in the fitted model fail
this test the model can be refitted using a larger basis expansion, say
by doubling the value of \emph{k} (the rank) used to fit the original.
Note also that a smooth might fail this test whilst using fewer
effective degrees of freedom than the maximum possible for the dimension
of basis used. This may happen when the true function lies at the upper
limit of the set of functions encompassed by the size of basis used.
Additionally, a basis of size \(2k\) encompasses a richer space of
functions of a given complexity than a basis of size \(k\) (Wood, 2017);
increasing the basis dimension used to fit the model may unlock this
additional function space resulting in a better fitting model whilst
using a similar number of effective degrees of freedom.

@@

With the larger basis dimension and accounting for the non-constant
variance of the observations via weights, the model fitted using REML is
indistinguishable from that obtained using GCV (Figure
\ref{fig:plot-fitted-models}b). The trace of the REML score for this
model shows a pronounced minimum at a much smaller value of \(\lambda\)
than the original REML fit (Figure
\ref{fig:trace-smoothness-parameters}), indicating that a more wiggly
trend provides a better fit to the Braya-Sø time series. This example
illustrates that some care and understanding of the underlying
principles of GAMs is required to diagnose potential issues with the
estimated model. After standard modelling choices (size of basis to use,
correct selection of response distribution and link function, etc.) are
checked, it often pays to think carefully about the properties of the
data and ensure that the assumptions of the model are met. Here, despite
increasing the basis size, it was the failure to appreciate the
magnitude of the effect of the non-constant variance that lead to the
initially poor fit and the problems associated with the estimation of
the CAR(1) process. I return to the issue of why the GAM plus CAR(1)
model encountered problems during fitting later
(\protect\hyperlink{identifiability}{see section Residual
autocorrelation and model identification}).

@@

If we want to ask whether either of the estimated trends is
statistically interesting or proceed to identify periods of significant
change, we must address the issue of uncertainty in the estimated model.
What uncertainty is associated with the trend estimates? One way to
visualise this is through a 1 - \(\alpha\) confidence interval around
the fitted trend, where \(\alpha\) is say 0.05 leading to a 95\%
interval. A 95\% interval would be drawn at
\(\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))\), with
\(m_{1-\alpha} = 1.96\), the 0.95 probability quantile of a standard
normal distribution\footnote{The 0.95 probability quantile of the
  \emph{t} distribution may be used instead, which will account for
  estimation of \(\sigma\), the variance of the data. However, given the
  number of observations, and hence residual degrees of freedom, needed
  to motivate fitting GAMs, differences between intervals computed using
  extreme quantiles of the standard normal or the \emph{t} distribution
  will be tiny.}, and \(\text{SE}(\hat{y}_t)\) is the standard error of
the estimated trend at time \(x_t\). This type of confidence interval
would normally be described as \emph{pointwise}; the coverage properties
of the interval being correct for a single point on the fitted trend,
but, if we were to consider additional points on the trend, the coverage
would logically be lower than 1 - \(\alpha\). This is the traditional
frequentist interpretation of a confidence interval. However, the GAM
described here has a Bayesian interpretation (Kimeldorf and Wahba, 1970;
Silverman, 1985; Wahba, 1983, 1990) and therefore the typical
frequentist interpretation does not apply. Nychka (1988) investigated
the properties of a confidence interval created as described above using
standard errors derived from the Bayesian posterior covariance matrix
for the estimated mode parameters. Such intervals have the interesting
property that they have good \emph{across-the-function} coverage when
considered from a frequentist perspective. This means that, when
averaged over the range of the function, the Bayesian credible intervals
shown in Figure \ref{fig:plot-fitted-models} have close to the expected
95\% coverage. However, to achieve this some parts of the function may
have more or less than 95\%-coverage. Marra and Wood (2012) recently
explained Nychka's (1988) surprising results and extended them to the
case of generalized models (non-Gaussian responses).

@@

Figure \ref{fig:posterior-simulation} shows 20 random draws from the
posterior distributions of the GAMs fitted to the Small Water and
Braya-Sø data sets. In the early period of the δ\textsuperscript{15}N
time series many of the posterior simulations exhibit short periods of
increasing and decreasing trend, balancing out to the relatively flat
trend estimated by the GAM (Fig.~\ref{fig:posterior-simulation}a).
Reflecting this uncertainty, we might expect relatively wide
simultaneous intervals during this period in order to contain the vast
majority of the simulated trends. Conversely, the decreasing
δ\textsuperscript{15}N trend starting at \textasciitilde{}1945 is
consistently reproduced in the posterior simulations, suggesting that
this feature of the time series is both real and statistically
significant, and that the rate of change in δ\textsuperscript{15}N is
relatively precisely estimated. We see a similar pattern in
Figure~\ref{fig:posterior-simulation}b for the Braya-Sø record; the
large peak in \uk{} at \textasciitilde{}250CE and the strong decline at
\textasciitilde{}1200CE are well defined in the posterior simulations,
whereas most of the localised trends that are smaller magnitude changes
in \(y_t\) are associated with posterior simulations that are less well
constrained with the ends of the record in particular showing
considerable variation in the strength, timing, and even sign of
simulated trends, reflecting the greater uncertainty in estimated trend
during these periods. For the random draws illustrated in Figure
\ref{fig:posterior-simulation}, a (1 - \(\alpha\))100\% simultaneous
interval should contain the entire function for on average 19 of the 20
draws.

@@

Simultaneous intervals computed using the method described are show in
Figure \ref{fig:compare-intervals} alongside the
\emph{across-the-function} confidence intervals for the trends fitted to
both example data sets. As expected, the simultaneous interval is
somewhat wider than the \emph{across-the-function} interval. The
critical value \(m_{1-\alpha}\) for the simultaneous interval of the
estimated trend in δ\textsuperscript{15}N is 3.07, whilst the same value
for the \uk{} series is 3.43, leading to intervals that are
approximately ±50\% and ±75\% wider than the equivalent
across-the-function intervals.

@@

Why might this be so? Autocorrelation is the tendency for a large
(small) value of \(y_t\) at time \(x_t\) to be followed by a likewise
large (small) value at time \(x_{t+1}\). This leads to runs of values
that are consistently greater (less) than the overall mean. Short runs
would indicate weaker autocorrelation whilst longer runs are associated
with stronger autocorrelation, and long runs of values greater (less)
than the mean would be evident as non-linear trends in the time series.
As a result, a wiggly trend and an autocorrelation function with large
\(\phi\) are two ways to describe the same pattern of values in a time
series, and, without any further information to constrain either of
these, the model is unable to distinguish both components uniquely.

@@

In the world of machine learning, Gaussian processes (Golding and Purse,
2016; Rasmussen and Williams, 2006) are a widely-used method for fitting
smooth non-parametric regression models. A Gaussian process is a
distribution over all possible smooth functions \(f(x)\). In the field
of spatial statistics, Gaussian processes are known by the name
\emph{kriging}.

@@

Each of the spline types that I have discussed so far shares a common
feature; the degree of wiggliness over the time series is fixed due to
the use of a single smoothness parameter, \(\lambda\). The definition of
wiggliness, as the integrated squared second derivative of the spline,
ensures that the fitted smoother does not jump about wildly. This
assumes that the data themselves are well described by a smoothly
varying trend. If we anticipate abrupt change or step-like responses to
environmental forcing this underlying assumption of the GAM would
suggest that the method is ill-suited to modelling palaeo time series in
which such features are evident or expected.

@@

Thus far, the trend models that I have described and illustrated assumed
that the time covariate (\(x_t\)) was fixed and known. In both examples,
and generally for most palaeoecological records, this assumption is
violated. Unless the record is annually laminated, assigning an age to a
sediment interval requires the development of an age model from
observations of the relationship between depth down the sediment core
and estimates of the age of the sample arrived at using any of a number
of techniques, for example \textsuperscript{210}Pb or
\textsuperscript{14}C radiometric dating. This age-depth relationship is
itself uncertain, usually being derived from a mathematical or
statistical model applied to point age estimates (e.g.~ Blaauw and
Heegaard, 2012). Incorporating this additional component of uncertainty
complicates the estimation of statistical models from
palaeoenvironmental data. In this section I illustrate a simulation
based approach to quantify and account for age-model uncertainty as part
of the trend estimation using a GAM (see Anchukaitis and Tierney (2013)
for a similar, non-GAM related idea).

@@

The ideas expressed in this paper are the result of many fruitful
conversations with colleagues past and present at the Environmental
Change Research Centre, UCL, and the University of Regina. In particular
I am indebted to Helen Bennion, Rick Battarbee, and Peter Leavitt for
their collaborations on projects over many years, and to David Miller,
Eric Pedersen, and Noam Ross, my GAM workshop partners in crime. Without
Simon Wood's \emph{mgcv} software and his research on GAMs, the
application of these models to palaeo time series would not be as
straight forward. This work was supported by a Natural Sciences and
Engineering Council of Canada (NSERC) Discovery Grant to the author
(RGPIN-2014-04032).
